---
title: "Data reading and storing with readr and tibble"
engine: knitr
---

```{r}
#| class-output: outputcode
#| eval: true
#| include: false
library(tidyverse)
```

In most data analytics projects, the first step is to load data into R from an external file.  The `readr` package is part of the `tidyverse` and was created to read in large, flat data files (e.g. csv, text) to R. The package is built on similar base R functions but has made them much more efficient. It also interacts consistently with other `tidyverse` packages. The class to store data.frames in the tidyverse is `tibble`. You can think of a tibble object as a (modern) data.frame.

## readr functions

There are several `readr` functions to choose from. These are:
    
* **read_csv()** - reads comma delimited files.
* **read_tsv()** - reads tab delimited files.
* **read_delim()** - reads in files with any delimiter.
* **read_table()** - reads in fixed width files where columns are separated by white space.

The function used will depend on the type of file that is being read into R.

## How to use read_csv()

A typical call to `read_csv()` has the form:
```{r}
#| class-output: outputcode
#| eval: false
tibble_name <- read_csv(file = "filepath/filename.csv")
```

The `file = ` argument is used to specify the file location (filepath), and the name of the file being read in (filename.csv). The function returns a tibble which can be stored with a name of your choice.

The `col_names` argument of the `read_csv()` function can be `TRUE` or `FALSE` or a character vector of column names. If `TRUE` (the default) the first row of input in the csv file will be used as the column names. If `FALSE` column names will be generated automatically (`X1`, `X2`, etc.). If a character vector, these will be used as the column names. By default, `read_csv()` reads in a csv file line-by-line. It will also read in the first few rows of the data and try to figure out the data type/class of each column. Sometimes it can make mistakes when doing this and often it is a good idea to explicitly specify the column types. This is achieved using the `col_types` argument. For example if a file has 10 columns, 8 of which are character and 2 of which are integer we can specify:

```{r}
#| class-output: outputcode
#| eval: false
my_dat <- read_csv(
    "data/filename.csv",
    col_types = "ccicccccci"
)
```

The `col_types = c` means that column stores character values. The `col_types = i` means that column stores integer values.

## Checking the data

It is good practice to check that the data have read in correctly to R. Use the `head()`, `tail()`, `summarise()`, `str()`, `dim()` commands to quickly check for any anomalies.

**NOTE**: To read in Excel files, another package (`readxl`) can be downloaded and used to achieve this.


## Additional resources

* `tibble` package: [link](https://tibble.tidyverse.org)

* `readr` package: [link](https://readr.tidyverse.org)

* Cheat sheet: [link](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf)

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 02](/weeks/week-2.qmd) ⏎
:::