---
title: "Load"
engine: knitr
---

<br>
As we have seen, the loading step is where transformed data goes into the target destination like a database, data warehouse, or storage system. Specific procedures ensure correct mapping to the target schema.

In this section, we will focus on the exportation of the data into an `.csv` file or an `.rds` file. You can also populate a SQL database, but it is outside the scope of the material. Please refer to the documentation of the `DBI` package for more information.

## Export as a csv file

A `.csv` file is a plain text file format used for storing tabular data. It is a widely used. Each line represents a row of data, and values within each row are separated by commas. CSV files are versatile and can be easily created, edited, and read using a wide range of software applications, including spreadsheet programs and text editors. It is a popular choice for data exchange and data storage as you can export the data using R and use the file with various other software.

::: {.callout-note appearance="simple"}

## Example

Using the dataset created in the previous section, we can save it as a `.csv` file using:
```{r}
#| eval: false
#| class-output: outputcode
write_csv(dataset_join, "./path/lord_of_the_rings.csv")
```

:::

We can load `.csv` files using the function `read_csv()`.

## Export as a rds file

An `.rds` file is a binary file format used in R for storing data objects. It encapsulates data and preserves the structure of R objects, making it a convenient choice for saving and sharing data in a compact and platform-independent (OS) manner. RDS files are often employed when the exact structure and content of an R object need to be preserved for future use or sharing with others. They provide a straightforward way to store data for analysis or to transport data between R sessions, but they can only be loaded in R...

::: {.callout-note appearance="simple"}

## Example

Using the dataset created in the previous section, we can save it as a `.rds` file using:
```{r}
#| eval: false
#| class-output: outputcode
write_rds(dataset_join, "./path/lord_of_the_rings.rds")
```

:::

We can load `.rds` files using the function `read_rds()`.


## Additional resources

* An [introduction](https://dbi.r-dbi.org/articles/dbi) to DBI.

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 03](/weeks/week-3.qmd) ⏎
:::