---
title: "Types and Classes in R"
engine: knitr
---

## R types

Types refer to the basic built-in data types that R understands natively. The most common types are:

- `logical` (a vector of logical (`TRUE/FALSE`) values)
- `double` (a vector of real numbers, i.e. have decimal places)
- `character` (a vector of character values)
- `closure` (for functions)
- `list` (for list)

You can access to the type of an object using the `typeof` function.

## Vectors definition

Vectors are a sequence of values of the same data types. There are six basic vector types: `logical`, `integer`, `double`, `complex`, `character` and `raw`.

Examples of vectors:
```{r vectors}
#| eval: false
double_vector <- c(1, 2, 3)
integer_vector <- c(1L, 2L, 3L)  # Note the difference with the previous vector.
character_vector <- c("apple", "banana", "peach")
```

Vector is the most basic data type. Single numbers or strings are vectors of length 1. Note that it is possible to define vectors with length zero.

## Lists

Lists are generic data structures that allows you to store different types of data objects. The *elements* of a list can be any data structure (even other lists).

Examples of lists:
```{r lists}
#| eval: false
l <- list(1, FALSE, "hello")

l <- list(
    a = c(1, 2, 3),
    b = "hello",
    c = c(TRUE, FALSE)
)
```

## R classes


## Additional resources

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 01](/weeks/week-1.qmd) ⏎
:::