---
title: "Objects in R"
engine: knitr
---

<br>
The entities manipulated by R are formally referred to as **objects**.

## Intrinsic attributes

An object has two intrinsic attributes: its **mode** and its **length**. The **mode** refers to the type of the object, while the **length** represents the number of elements the object can have.

### Mode

The mode of an object is the basic type of its fundamental components. Vectors must have all their values (or components) of the same mode. Vectors are **atomic** structures as their components have the same type (numeric, complex, logical, character, or raw). Vectors have thus one of the five types without ambiguity.

```{r}
#| class-output: outputcode
#| eval: true
A <- c(1, 2, 3)
mode(A)
```

::: {.callout-note appearance="simple"}
Note that the terms **mode** and **type** can be used (almost) interchangeably.
:::

More complex objects, such as lists or functions, are **recursive** structure (and not atomic). Their individual components may be of any mode (even list or function). The mode of a list is `list` and the mode of a function is `function`.

### Length

The length of an object defines how many elements the object contains. It can be accessed (and also modify) using the `length` function.

::: {.callout-note appearance="simple"}

## Example

First, we create a simple vector of length $10$. 

```{r}
#| class-output: outputcode
#| eval: true
A <- 1:10
print(A)
```

We then set the length of the vector to be $3$. It cuts all the elements that are in position greater than $3$.
```{r}
#| class-output: outputcode
#| eval: true
length(A) <- 3
print(A)
```

Finally, we want this vector to have a length of $6$. R will fill the position $4$ to $6$ of this vector with `NA`. Be careful that it will not recover the previous values, once the length is changed, the values are removed.
```{r}
#| class-output: outputcode
#| eval: true
length(A) <- 6
print(A)
```

:::

### Empty objects

In R, there is no problem to define empty objects with a particular mode. To create such a vector, the functions are `integer()`, `complex()`, `logical()`, `character()` and `raw()`. You can also define lists using empty vectors. The length of an empty vector is $0$.

```{r}
#| class-output: outputcode
#| eval: true
null_vector <- integer()  # Vector of integer of length zero.
```

```{r}
#| class-output: outputcode
#| eval: true
mode(null_vector)
```

```{r}
#| class-output: outputcode
#| eval: true
length(null_vector)
```



## R classes

A class represents the structure of an object. All objects have a class reported by the `class` function. It provides essential information about how an object is organized and interpreted by R and how it can be manipulated. Each class has its own set of methods and functions. It permits an object-oriented programming approach. You can define your own classes using the `class` method and define methods for this class using `name_function.name_class`. This feature is particularly convenient to override existing methods.

::: {.callout-note appearance="simple"}

## Example

To create a new class for a simple vector and override the `print` function.

```{r}
#| class-output: outputcode
#| eval: true
A <- c(1, 2, 3)
```

```{r}
#| class-output: outputcode
#| eval: true
typeof(A)  # type of A
class(A)  # class of A
print(A)  # print A
```

```{r}
#| class-output: outputcode
#| eval: true
class(A) <- "my_class"

print.my_class <- function(obj) {
    print(paste("Vector: ", obj))
}
```

```{r}
#| class-output: outputcode
#| eval: true
typeof(A)  # type of A
class(A)  # class of A
print(A)  # print A
```

So, the type of the object did not change, but we have defined our own class with a particular `print` function.

:::

## Additional resources

- R manual by the R Core Team, [Section 3 on objects](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Objects).


<br><br>

::: {style="font-size: 0.875em;"}
[Back](/weeks/week-1.qmd) ⏎
:::