---
title: "Logical operators"
engine: knitr
---

## Logical operators

Logical operators are used to compare objects. Every logical operator returns a
logical (boolean) vector of `TRUE`/`FALSE` values. We can then use these `TRUE`
and `FALSE` values to create subsets of our data. The primary logical operators
that we can use in R are

| Operator | Meaning                        |
|:--------:|:-------------------------------|
|   `==`   | equality                       |
|   `!=`   | non-equality                   |
|   `<`    | less than                      |
|   `>`    | greater than                   |
|   `<=`   | less than or equal to          |
|   `>=`   | greater than or equal to       |
|   `!`    | returns opposite boolean value |


::: {.callout-note appearance="simple"}

## Example

For these examples, let's set up a numeric vector `x`:

```{r}
#| class-output: outputcode
#| eval: true
x <- c(1, 3.5, 7.9, 10, 12.3, NA, 15.8)
```

Now, let's use some logical operators:

```{r}
#| class-output: outputcode
#| eval: true
x == 3.5
```

```{r}
#| class-output: outputcode
#| eval: true
x != 3.5
```

Notice that the `!` operator changes `TRUE` to `FALSE` and vice-versa.

If we want to identify where we have values $> 3.5$, we use:

```{r}
#| class-output: outputcode
#| eval: true
x > 3.5
```

If we use the `!` operator:

```{r}
#| class-output: outputcode
#| eval: true
!(x > 3.5)
```

This is equivalent to getting the values $\leq 3.5$:
```{r}
#| class-output: outputcode
#| eval: true
x <= 3.5
```

:::

## The OR and AND operators

Two other very useful operators are the `|` (the OR operator) and the `&` (the
AND operator). These are used to combine two or more logical operations. The `|`
(OR) operator returns `TRUE` if any of the conditions are met. The `&` (AND)
operator returns `TRUE` if both of the conditions are met. There are other
functions such as `any()`, `all()` and `%in%` that you can use to apply OR and
AND operators on logical vectors.

The table for the OR operator is

|   `x`   |   `y`   | <code>x \| y</code> |
|:-------:|:-------:|:-------------------:|
| `TRUE`  | `TRUE`  |       `TRUE`        |
| `FALSE` | `TRUE`  |       `TRUE`        |
| `TRUE`  | `FALSE` |       `TRUE`        |
| `FALSE` | `FALSE` |       `FALSE`       |

::: {.callout-note appearance="simple"}

## Example

Say we want to return `TRUE` if the value of `x` is $> 7.9$ OR $< 3.5$:

```{r}
#| eval: true
#| class-output: outputcode
x > 7.9 | x < 3.5
```

:::


The table for the AND operator is

|   `x`   |   `y`   | <code>x & y</code> |
|:-------:|:-------:|:------------------:|
| `TRUE`  | `TRUE`  |       `TRUE`       |
| `FALSE` | `TRUE`  |      `FALSE`       |
| `TRUE`  | `FALSE` |      `FALSE`       |
| `FALSE` | `FALSE` |      `FALSE`       |

::: {.callout-note appearance="simple"}

## Example

Say we want to return `TRUE` if the value of `x` is $< 7.9$ AND $> 3.5$:

```{r}
#| eval: true
#| class-output: outputcode
x < 7.9 & x > 3.5
```

:::

The `|` and `&` operators perform elementwise comparisons. So, they return a vector. To evaluate the statement from left to right, we use the `||` and `&&` operators. These operators can not be applied on vectors with more than one element. This is the appropriate form of the OR and AND operators in conditional statements.

## Additional resources

- The Logic documentation page ([?base::Logic](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Logic.html))

<br><br>

::: {style="font-size: 0.875em;"}
[Back](/weeks/week-1.qmd) ⏎
:::