---
title: "Vectors"
engine: knitr
---

## Creating a vector

A vector is created using the `c()` function. To create a numeric vector with the values 1, 2, 3 and 4, we can use:

```{r vec_create}
#| eval: true
#| class-output: outputcode
vec <- c(1, 2, 3, 4)
vec  # Print the contents
```

## Length and type of a vector

The `length()` function is used to count the number of elements within a vector.

```{r vec_length}
#| eval: true
#| class-output: outputcode
vec <- c(10.4, 5.6, 3.1, 6.4, 21.7)  # Numeric vector with 5 elements
length(vec)
```

To check the type of a vector, you can use:

```{r vec_type}
#| eval: true
#| class-output: outputcode
typeof(vec)
```

## Combining vectors

Vectors can be combined using the `c()` command. For example, to combine twice the `vec` vector:

```{r vec_combine}
#| eval: true
#| class-output: outputcode
c(vec, vec)
```

Be careful, if you try to combine vectors with different type, the type of the resulting vector will be the most general one. For example, here, the numbers are converted to characters.

```{r vec_combine_type}
#| eval: true
#| class-output: outputcode
c(vec, "Mary")
```

## Indexing vectors

Accessing to specific element(s) of a vector is possible using the `[]` operator. For example, `vec[2]` will return the second element of the vector `vec`

```{r vec_index}
#| eval: true
#| class-output: outputcode
vec[2]
```

Specifying negative values in the indices will drop that element(s).

```{r vec_index_neg}
#| eval: true
#| class-output: outputcode
vec[-2]
```

Multiple elements can be accessed by supplying a vector of indices. To access the first and third elements of `vec`:

```{r vec_index_multi}
#| eval: true
#| class-output: outputcode
vec[c(1, 3)]
```

To access consecutive indices, you can use `:`. For example, to access the first, second and third elements of `vec`:

```{r vec_index_multi_consecutive}
#| eval: true
#| class-output: outputcode
vec[1:3]
```

Logical operators can be used for subsetting vectors. If we want to work with just the components of `vec` that are larger than 6:

```{r vec_index_logic}
#| eval: true
#| class-output: outputcode
mask <- vec > 6
vec[mask]
```

Note that it could be done in one line using `vec[vec > 6]`. R returns the values in `vec` where a `TRUE` is recorded for `vec > 6`.

## Operations

Usual mathematical operators can be carried on vectors:

-   Addition: `+`
-   Subtraction: `-`
-   Multiplication: `*`
-   Division: `/`
-   Euclidean division: `%%` and `%/%`
-   Exponentiation: `^`

Computations are (typically) performed **element-wise**. Recall the vector `vec`:

```{r}
#| eval: true
#| class-output: outputcode
vec <- c(10.4, 5.6, 3.1, 6.4, 21.7)
```

If we calculate `vec^2`, we get:

```{r expo}
#| eval: true
#| class-output: outputcode
vec^2
```
i.e. R takes each elements of `vec` and calculates `value^2`.

If you are unsure if the calculations are element-wise, you should check on some elements. Other examples of common operations are `log()`, `exp()`, `sin()`, `cos()`, etc.

```{r vec_op}
#| eval: true
#| class-output: outputcode
log(vec)
exp(vec)
sin(vec)
```

## Vectors of the same length

We can add, subtract, multiply or divide vectors of the same length easily. In this case, R takes the values in the same locations (e.g. the first element of each vector) and performs the computations. For example, to add two vectors of the same length:

```{r two_vec_op}
#| eval: true
#| class-output: outputcode
vec_1 <- c(1, 3, 5, 7, 9)
vec_2 <- c(0, 2, 4, 6, 8)

vec_1 + vec_2
```


## Vector of different lengths

To carry out an operation on two vectors of different lengths, the shorter vector is **recycled** to match the length of the longer one. Let's look at an example to understand the concept.

```{r recycled_vec}
#| eval: true
#| class-output: outputcode
vec_1 <- c(1, 3, 5, 7, 9)
vec_2 <- c(0, 2)

vec_1 * vec_2
```

Let's dive into the different elements of the resulting vector.

-   The first element is `1 * 0 = 0`.
-   The second element is `3 * 2 = 6`.
-   The third element is `5 * 0 = 0`.
-   The fourth element is `7 * 2 = 14`.
-   The fifth element is `9 * 0 = 0`.

The version of `vec_2` that was actually multiplied to `vec_1` is therefore:

```{r}
#| eval: false
c(0, 2, 0, 2, 0)
```

## Additional resources


<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 01](/weeks/week-1.qmd) ⏎
:::