---
title: "If-Else statement"
engine: knitr
---


The `if-else` combination is extremely common in R and in many other programming languages. These statements allow you to execute chunks of code if a particular condition is met (or not). We will look at `if`, `else` and `else if` statements.

## if statements

The basic format of an `if` statement is:

```{r}
#| class-output: outputcode
#| eval: false
if (condition) {
    # Execute this code
}
    
# Otherwise continue with rest of code
```

The `condition` must be a logical operator, i.e. return only a TRUE or FALSE value. If the `condition` is TRUE, the code between the `{ }` will be executed.  If the `condition` is FALSE, the code between the `{ }` will be skipped and the next line(s) of code executed.

For example, a random number between $0$ and $100$ is generated and stored in `x`.  **Pass** will be printed if `x` is $>50$. The code is as follows:

```{r}
#| class-output: outputcode
#| eval: true
x <- sample(1:100, 1)
if (x > 50) {
    print("Pass")
}

```

## else statements

The latter code will do nothing if the `condition` returns FALSE. If you want something specific to happen, an `else` statement is needed. This has the general form:
```{r}
#| class-output: outputcode
#| eval: false
if (condition) {
    # Execute this code
} else {
    # Execute this code instead
}
```
    

For example, a random number between $0$ and $100$ is generated and stored in `x`. **Pass** will be printed if `x` is $>50$. **Fail** will be printed otherwise.

The code is as follows:
```{r}
#| class-output: outputcode
#| eval: true
x <- sample(1:100, 1)

if (x > 50) {
    print("Pass")
} else {
    print("Fail")
}

```


## if, else if, else statements

You can have a series of tests by starting with an `if` statement and following it with several `else if` statements. This has the general form:

```{r}
#| class-output: outputcode
#| eval: false
if (condition_1) {
    # Execute code chunk 1
} else if (condition_2) {
    # Execute code chunk 2
} else {
    # Execute this code
}
```


For example, a random number between $0$ and $100$ is generated and stored in `x`.  **Pass** will be printed if `x` is $>50$. **Check** will be printed if `x` is $>30$ and $\leq 50$. **Fail** will be printed otherwise.

The code is as follows:
```{r}
#| class-output: outputcode
#| eval: true
x <- sample(1:100, 1)
if (x > 50) {
    print("Pass")
} else if (x > 30 & x <= 50) {
    print("Check")
} else {
    print("Fail")
}
```


## Additional resource


<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 01](/weeks/week-1.qmd) ⏎
:::