---
title: "Loops"
engine: knitr
---

<br>
Loops are chunks of code that run repeatedly until a stopping criterion is reached. There are lots of different types of loops that are available, e.g. `while` loops, `repeat` loops, and `for` loops. We will focus on `for` loops as these are the most common type of loop you will need when programming in R.

::: {.callout-note appearance="simple"}

It can be better for code efficiency to avoid using loops, using vectorisation for example, but often a loop works just fine!

:::


## Structure of for loops

To loop/iterate through a certain number of repetitions, or carry out the same thing to lots of inputs (e.g. over all columns of a data frame), a `for` loop is used. In R, `for` loops take an **iterator variable** and assign it successive values from a sequence or vector. The general structure of a `for` loop is:

```{r}
#| class-output: outputcode
#| eval: false
for (iterator_variable in some_sequence) {
    # do something
}
```

Often we use `i` to denote the iterator variable name but any name can be used. Each time the loop is run, `i` then changes value as given by the numbers in `some_sequence`. Typically `i` increments by $1$ (or some integer), but not always.

::: {.callout-note appearance="simple"}

## Example

First, we consider a simple example where we want to print the numbers from $1$ to $10$.

```{r}
#| class-output: outputcode
#| eval: true
for (i in 1:10) {
    print(i)
}
```

Here `i` is the iterator variable and the sequence $1, 2, 3, \dots, 10$ is the vector of values `i` will take on. First `i` is given the value $1$, and the code between the curly brackets is executed. Here $1$ is printed to the screen. Once the code is executed, `i` is then incremented to the next value in the sequence $1, 2, 3, \dots, 10$, i.e. the number $2$. The code is executed again and $2$ is printed to the screen. The loop repeats until it reaches the final value in the sequence, $10$, executes the code and then exits the loop. 

:::


::: {.callout-note appearance="simple"}

## Example

Consider a vector of characters named `char_vec`:
```{r}
#| class-output: outputcode
#| eval: true
char_vec <- c("a", "b", "c", "d")
```

To print the contents of `char_vec` there are several options. The following three loops all have the same behaviour.

1. Iterate directly through `char_vec` as the previous example:
```{r}
#| class-output: outputcode
#| eval: true
# Loop 1
for (i in char_vec) {
    print(i)
}
```
    

2. Define a sequence of indices and access each element of `char_vec` using `[ ]`:
```{r}
#| class-output: outputcode
#| eval: true
# Loop 2
for (i in 1:4) {
    print(char_vec[i])
}
```

3. Use `seq_along()` to automatically create a sequence of indices of the correct length:
```{r}
#| class-output: outputcode
#| eval: true
# Loop 3
for (i in seq_along(char_vec)) {
    print(char_vec[i])
}
```

:::

## Storing outputs

When using `for` loops we often want to store the results of the code we are executing in the loop. Typically, the results will be stored in specific locations in a vector, data frame, or list depending on the value of `i`. 

::: {.callout-note appearance="simple"}

## Example

Let's set up a list with $3$ elements, each of which is a numeric vector:

```{r}
#| class-output: outputcode
#| eval: true
data_list <- list(
    el_1 = c(1, 2, 3, 4, 5),
    el_2 = c(3.4, 2.3, 3.1, 4),
    el_3 = c(19.3, 12.5, 6.1)
)
```

Say we want to calculate and store the mean of each of the vectors in `data_list`. A `for` loop can be used to do this:

```{r}
#| class-output: outputcode
#| eval: true
# Set up an empty numeric vector to store the means. 
mean_vec <- vector(
    mode = "numeric",
    length = length(data_list)
)

# Run the for loop 
for (i in seq_along(data_list)) {
    mean_vec[i] <- mean(data_list[[i]]) 
}

mean_vec
```

:::

There are 3 steps in this process:

1. **Output** - set up an empty object to store the results.  
2. **Sequence** - set up the iterator variable and the sequencing vector.  
3. **Body** - write the code to execute.
    
## Nested loops

`for` loops can also be **nested**, i.e. we can have multiple loops iterating over different indices. For example, we might want to iterate over the rows and columns of a data frame.

::: {.callout-note appearance="simple"}

## Example

A nested loop is constructed as follows:
```{r}
#| class-output: outputcode
#| eval: true
my_data_frame <- data.frame(var1 = 1:6, var2 = 8:13)
for (i in 1:nrow(my_data_frame)) {
    for (j in 1:ncol(my_data_frame)) {
        print(my_data_frame[i, j])
    }   
}
```

:::

## Implicit looping

R has an in-built family of functions which implement looping in a compact and efficient way. The main functions in this family are:  
  
* `lapply()` - loop over a list and evaluate a function on each element.
* `sapply()` - same as `lapply()` by tries to simplify the result by returning a vector, matrix or array.
* `tapply()` - apply a function over subsets of a vector.
* `apply()` - apply a function over the margins of an array.
* `mapply()` - a multivariate version of `lapply()`.

::: {.callout-note appearance="simple"}

## Example

To calculate the mean for each element of the data frame `data_list` without using a `for` loop:

```{r}
#| class-output: outputcode
#| eval: true
data_list <- list(
    el_1 = c(1, 2, 3, 4, 5),
    el_2 = c(3.4, 2.3, 3.1, 4),
    el_3 = c(19.3, 12.5, 6.1)
)
sapply(data_list, mean)
```

:::

## Additional resource

- R manual by the R Core Team, [Section 9 on repetitive execution](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Repetitive-execution).

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 01](/weeks/week-1.qmd) ⏎
:::