---
title: "Loops"
engine: knitr
---

## What are loops?

Loops are chunks of code that run repeatedly until a stopping criterion is reached. There are lots of different types of loops that are available, e.g. `while` loops, `repeat` loops, and `for` loops. We will focus on `for` loops as these are the most common type of loop you will need when programming in R.

**NOTE**: it can be better for code efficiency to avoid using loops but often a loop works just fine!

## Structure of for loops

To loop/iterate through a certain number of repetitions, or carry out the same thing to lots of inputs (e.g. over all columns of a data frame), a `for` loop is used. In R `for` loops take an *iterator variable* and assign it successive values from a sequence or vector. The general structure of a `for` loop is:

```{r}
#| class-output: outputcode
#| eval: false
for (iterator_variable in some_vector) {
    # do something
}
```

Often we use `i` to denote the iterator variable name but any name can be used. Each time the loop is run, `i` then changes value as given by the numbers in `some_vector`. Typically `i` increments by 1 (or some integer), but not always.

## A simple for loop

In the simplest example:
```{r}
#| class-output: outputcode
#| eval: true
for (i in 1:10) {
    print(i)
}
```

Here `i` is the iterator variable and the sequence 1, 2, 3, ..., 10 is the vector of values `i` will take on. First `i` is given the value 1, and the code between the curly brackets is executed. Here 1 is printed to the screen.  Once the code is executed, `i` is then incremented to the next value in the sequence 1, 2, 3, ..., 10, i.e. the number 2. The code is executed again and 2 is printed to the screen. The loop repeats until it reaches the final value in the sequence, 10, executes the code and then exits the loop. 

## Iterating

The vector of values to iterate through can have many structures. To print even numbers between 1 and 10 use:  
```{r}
#| class-output: outputcode
#| eval: true
steps <- c(2, 4, 6, 8, 10)
for (i in steps) {
    print(i)
}
```

Next say we have a character vector called `char_vec`:
```{r}
#| class-output: outputcode
#| eval: true
char_vec <- c("a", "b", "c", "d")
```

To print the contents of `char_vec` there are several options. The following three loops all have the same behaviour.

Iterate directly through `char_vec`:
```{r}
#| class-output: outputcode
#| eval: true
# Loop 1
for (i in char_vec) {
    print(i)
}
```
    
Define a sequence of indices and access each element of `char_vec` using `[ ]`:
```{r}
#| class-output: outputcode
#| eval: true
# Loop 2
for (i in 1:4) {
    print(char_vec[i])
}
```

Use `seq_along()` to automatically create a sequence of indices of the correct length:
```{r}
#| class-output: outputcode
#| eval: true
# Loop 3
for (i in seq_along(char_vec)) {
    print(char_vec[i])
}
```

## Storing outputs

When using `for` loops we often want to store the results of the code we are executing in the loop. Typically the results will be stored in specific locations in a vector, data frame, or list depending on the value of `i`. Let's set up a list with 3 elements, each of which is a numeric vector:

```{r}
#| class-output: outputcode
#| eval: true
dat.list <- list(
    el_1 = c(1, 2, 3, 4, 5),
    el_2 = c(3.4, 2.3, 3.1, 4),
    el_3 = c(19.3, 12.5, 6.1)
)
```

Say we want to calculate and store the mean of each of the vectors in `dat.list`. A `for` loop can be used to do this:

```{r}
#| class-output: outputcode
#| eval: true
# Set up an empty numeric vector to store the means. 
mean.vec <- vector(mode = "numeric", length = length(dat.list))

# Run the for loop 
for (i in seq_along(dat.list)) {
    mean.vec[i] <- mean(dat.list[[i]]) 
}

mean.vec
```
    
There are 3 steps in this process:

1. **Output** - set up an empty object to store the results.  
2. **Sequence** - set up the iterator variable and the sequencing vector.  
3. **Body** - write the code to execute.
    
## Nested loops

`for` loops can also be *nested*, i.e. we can have multiple loops iterating over different indices. For example, we might want to iterate over the rows and columns of a data frame.

A nested loop is constructed as follows:
```{r}
#| class-output: outputcode
#| eval: true
my.data.frame <- data.frame(var.1 = 1:6, var.2 = 8:13)

for (i in 1:nrow(my.data.frame)) {
    for (j in 1:ncol(my.data.frame)) {
        print(my.data.frame[i, j])
    }   
}
```

## Implicit looping

R has an in-built family of functions which implement looping in a compact and efficient way. The main functions in this family are:  
  
* `lapply()` - loop over a list and evaluate a function on each element
* `sapply()` - same as `lapply()` by tries to simplify the result
* `tapply()` - apply a function over subsets of a vector
* `apply()` - apply a function over the margins of an array
* `mapply()` - a multivariate version of `lapply()`

To calculate the mean for each element of the data frame `dat.list` without using a `for` loop:

```{r}
#| class-output: outputcode
#| eval: true
dat.list <- list(
    el_1 = c(1, 2, 3, 4, 5),
    el_2 = c(3.4, 2.3, 3.1, 4),
    el_3 = c(19.3, 12.5, 6.1)
)
lapply(dat.list, mean)
```


## Additional resource


<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 01](/weeks/week-1.qmd) ⏎
:::