---
title: "Data frames"
engine: knitr
---

<br>
A data frame is a data structure in R that represents tabular data, where rows and columns correspond to observations and variables, respectively. This struture is used for data manipulation, analysis, and modeling. A data frame can be thought of as a two-dimensional array-like structure with additional features that make it highly versatile for real-world data analysis tasks. The mode of a data frame is `list`, but its class is `data.frame`. 

## Creating a data frame

A data frame is created using the `data.frame()` function. It is a list with constrained:

1. The elements should consist of vectors (numeric, character, or logical), factors, numeric matrices, lists, or other data frames.

2. When matrices, lists, or data frames are used, they contribute variables to the new data frame corresponding to their columns, elements, or variables, respectively.

3. Variables represented as vectors within the data frame must share the same length, while matrix structures should have an identical number of rows.

::: {.callout-note appearance="simple"}

## Example

```{r}
#| class-output: outputcode
#| eval: true
# Creating a data frame
student_data <- data.frame(
    student_id = c(1, 2, 3, 4),
    name = c("Alice", "Bob", "Cathy", "David"),
    age = c(22, 23, 21, 24),
    grade = c("A", "B", "B", "C")
)

student_data
```
:::

## Indexing data frames

Data frames are lists. So, you can index data frames the same way you would index lists, using `[ ]`, `[[ ]]` or `$`. The `$` is particularly useful in the case of data frames because the variables are named.

::: {.callout-note appearance="simple"}

## Example

To select the name of the students in the data, we can use:

```{r}
#| class-output: outputcode
student_data$name
```

:::

## Appending to data frames

Adding a new variable (column) to a data frame is fairly simple as it works as a list. So, you can use position indexing (with `[[ ]]`) or name indexing (with `$`) to add a varible.

::: {.callout-note appearance="simple"}

## Example

To add a variable with the class of the student:
```{r}
#| class-output: outputcode
student_data$class <- c("Maths", "Stat", "Stat", "Maths")
student_data
```

:::

Adding a new observation (row) to a data frame is a bit more difficult as it consists to add new values to each components of the list (in the sense of data frame). To do, you will need to create a new data frame with the new observation(s) and bind the two data frames.

::: {.callout-note appearance="simple"}

## Example

To add another student:
```{r}
#| class-output: outputcode
new_student <- data.frame(
    student_id = 5,
    name = "Edward",
    age = 23,
    grade = "A",
    class = "Stat"
)
student_data <- rbind(student_data, new_student)
student_data
```

:::


## Additional resources

- R manual by the R Core Team, [Section 6 on data frames](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Data-frames).

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 01](/weeks/week-1.qmd) ⏎
:::