---
title: "Memory"
engine: knitr
---

<br>

In this section, we will give a general overview into how the memory is adressed in a computer and in R.

## Organisation of the RAM

The Random-Access Memory (RAM) is organised as an ordered sequence of boxes, each box can contain a binary digit: $0$ or $1$ (see @fig-ram). The information contained in a box, the smallest quantity of information that can be contained, is called a **bit** (for binary digit). Information is actually often organised in blocks of $8$ boxes.

Another unit has thus been introduced: the **byte**, which is worth $8$ bits. Note that each block is numbered; the number of a block is called its **memory address**. A memory address is thus an identifier, which designates a specific zone of memory where data (or instructions to run) can be read and stored. This identifier is usually an integer expressed in hexadecimal notation (base $16$).

::: {#fig-ram}

![](./figures/RAM.svg)

Schema of the RAM.

:::

## Accessing the memory

To access a given zone of memory, R uses what is called a **pointer**. A pointer is a variable containing a memory address (it points to the desired memory zone). R does that in a transparent way hidden from the user. At the address contained in a given pointer, we can find for example a data point. Remember that each data point has a specific type, such as integer, double, etc.

An integer is coded on $4$ bytes, a double on $8$ bytes, a character on $1$ byte, a logical on $4$ bytes and a complex on $16$ bytes to cite the most common variable types. This is true on a $32$ bit processor and on a $64$ bit processor.

```{r}
#| class-output: outputcode
#| eval: true
x <- 3L  # Create the value 3 of type integer.
```

So, what happen when the previous command is run? A memory slot of $32$ successive blocks ($4$ bytes of $8$ bits each) is allocated (or reserved) and a pointer is created containing the address of the first of these boxes. Actually, the pointer must not only contain the address of the variable `x` but also its type to know over how many boxes the variable is stored. The pointers are said to be **typed**. When a typed pointer is incremented, i.e., when we need to add one unit to the address it contains, it is not necessarily incremented by $1$, but by the size of the pointed type. For example, with integers, it will be incremented by $4$.


## Issues with integers

Since a (signed) integer is coded over $4$ bytes, i.e. $32$ bits, the largest integer that can be represented is $2147483647$. Indeed, if the first bit is reserved for the sign, there are $31$ remaining available boxes, or $2^{31}$ possible arrangements. Counting $0$, the largest available integer is this $2^{31} - 1$:
```{r}
#| class-output: outputcode
#| eval: true
as.integer(2^31 - 1)  
.Machine$integer.max
```

If you try to convert $2^{31}$ as integer, you will get an `NA`. You need a *double* variable to handle number larger than $2^{31}$.

```{r}
#| class-output: outputcode
#| eval: true
as.integer(2^31)  
is.double(2^31)
```

As the length of a vector (or matrix) is stored as a signed integer, this result has a very pratical application. You can not create a vector with more than $2^{31} - 1$ elements, whether on a $32$ bit or $64$ bit processor. 

::: {.callout-note appearance="simple"}

## Example

Regarding how the variables are stored, it can result in an overflow (or not).

```{r}
#| class-output: outputcode
#| eval: true
46360 * 46360  # 46360 is stored as a double.
46360L * 46360L  # 46360L is stored as an integer.
```

It might be tricky when combined with usual mathematical operations.

```{r}
#| class-output: outputcode
#| eval: true
sum(1:304)  # sum(1:304) is stored as an integer.
sum(1:304) * sum(1:304)  # is a product of integer.
46360^2  # 46360 is stored as a double
sum(1:304)^2  # ...
```

The last command, following what happened previously, should have result in an integer overflow warning. The exponent function, however, transforms its arguments into reals and returns a real number. 
:::

## Object size in R

We may expect that the size of the object `3L` to be $8$ bytes, but it is not actually the case.

```{r}
#| class-output: outputcode
#| eval: true
object.size(3L)  # on a 64 bit processor
```

In fact, in R, each object contains a header which takes up some space in the RAM: $24$ bytes on a $32$ bit processor and $40$ bytes on a $64$ bit processor. This header is used to save information of the created object: its type, its length, etc.

We can find out what kind of processor R is runnig using 
```{r}
#| class-output: outputcode
#| eval: true
.Machine$sizeof.pointer
```

The value $8$ is returned for a $64$ bit processor and the value $4$ for a $32$ bit processor. Nowadays, $32$ bit processor are not really used anymore.


## Total memory used by R

The total size of virtual memory allocated to R in a session includes:

* memory used to store the values of the objects;
* memory used to store the headers of the objects.

The information is accessible using the `gc()` functions. The variable `Ncells` represents the number of cells used for the headers and the variable `Vcells` the number of blocks for the values.

The total amount of memory available to R depends on several factors:

* the amount of RAM physically present on the computer;
* the amount of RAM already used by the operating system and the other softwares being run on that system (e.g. web browser);
* the type of the processor. On a $64$ bit processor, an address is coded over $64$ bits, so theoretically, the addressable memory is *limited* to $2^{64}$ bytes. In practice, it is usually limited by the processor's architeture.

So R allocates memory for the creation of (large) objects. When they are no longer used, these objects are cleared from memory. This process is called **garbage collection**. This process is usually automatic but you can force a garbage collection using the function `gc()`. When creating a large object in R, the memory reserved must be contiguous, i.e. it cannot be fragmented into multiple blocks. It is thus possible that there remains enough total memory for R, but no gap large enough to fit the data of a single large object.

## Some recommendations

An elementary understanding of memory management on a computer, in general, and in R, in particular, could be very useful to help identify the origin of memory-related issues.

* **Calculate the size of the object**: Before creating an object, we can calculate an approximation of its size. For example, imagine we want to create a matrix of real number of size $n \times p$. Since a real number uses up to $8$ bytes, this matrix will require $8np$ bytes. A square matrix of size $16000 \times 16000$ will roughly use $2$ Gb.
* **Remove useless objects**: If you are not able to create a large object, we can try to remove other useless large objects, using the function `rm()`, and to free up some memory using `gc()`. You can also close other running software.
* **Split the problem into multiple ones**: Returning to the matrix problem, you can split the matrix into several submatrices and find a way to perform your analysis on those, before combining the results (e.g. the MapReduce framework).

## Additional resources

* The Wikipedia RAM [page](https://en.wikipedia.org/wiki/Random-access_memory).

* MapReduce [page](https://www.databricks.com/glossary/mapreduce).

<br><br>

::: {style="font-size: 0.875em;"}
[Back](/weeks/week-7.qmd) ⏎
:::