---
title: "Simulation from a mixture of distributions"
engine: knitr
---

```{r}
#| eval: true
#| include: false
#| warning: false
#| class-output: outputcode
library(tidyverse)
```

<br>
A mixture of distributions refers to a statistical model that combines multiple probability distributions. This technique is used when a single probability distribution cannot adequately capture the underlying patterns in the data. Each of the underlying distributions has its own set of parameters.

## The composition method

Frequently, a single distribution is insufficient for accurately modelling the data at hand. For example, consider a dataset of heights that includes both men and women. In such a scenario, we would anticipate the data to exhibit a bimodal pattern with two distinct peaks, each corresponding to the mean height of the respective gender. To better represent such data, a mixture of distributions is preferred over a single distribution.

Assume that the probability density function $f$ is a convex mixture of probability density functions $f_k$:

$$X \sim f(x) = \sum_{k = 1}^K \pi_k f_k(x),$$

where for all $k = 1, \dots, K$, $\pi_k > 0$ and $\pi_1 + \dots + \pi_K = 1$.
We introduce an artificial auxiliary variable $Y \sim \pi$ such that $\pi_k = P(Y = k)$ and the conditional probability function of $X$ given $Y = k$ is $f_k(\cdot)$. Then, generating $X$ with the propability density function $f$ is equivalent to generate the vector $(X, Y)$. 

::: {.callout-note icon=false}
## Algorithm: the composition method

Input:

* $n$: length of the sample
* $f_k, k = 1, \dots, K$: probability density functions

Algorithm:

<div class="pseudocode">

**for** $i$ **from** $1$ **to** $n$ **do**

* generate $y_i \sim \pi(\cdot)$
* generate $x_i = f_{y_i}(\cdot)$

**endfor**
</div>

Output:

* Sample $(x_1, \dots, x_n)$ with probability density function $f$.
:::


::: {.callout-note appearance="simple"}

## Example: Simulation of a sample from a Gaussian mixture models.

We aim to generate a sample $x_1, \dots, x_n$ i.i.d. from 
$$f(x) = 0.5f_1(x) + 0.25f_2(x) + 0.25f_3(x),$$
where $f_1$ the pdf of $\mathcal{N}(0, 1)$, $f_2$ is the pdf of $\mathcal{N}(2, 2.25)$ and $f_3$ is the pdf of $\mathcal{N}(4, 2.25)$.

```{r}
#| class-output: outputcode
#| eval: true
set.seed(42)  # seed setting
n <- 10000  # number of samples   
probs <- c(0.5, 0.25, 0.25)  # probabilities to be in each mixture

# Generate the pi variable and each mixture separetely
pi <- sample(c(1, 2, 3), n, replace = TRUE, prob = probs)
f1 <- rnorm(n)
f2 <- rnorm(n, 2, 1.5)
f3 <- rnorm(n, 4, 1.5)

# Create the mixture
f <- (pi == 1) * f1 + (pi == 2) * f2 + (pi == 3) * f3
```

```{r}
#| class-output: outputcode
#| eval: true
set.seed(42)

density_estim <- density(f, from = -5, to = 10)
density_true <- (
    0.5 * dnorm(density_estim$x) +
    0.25 * dnorm(density_estim$x, 2, 1.5) +
    0.25 * dnorm(density_estim$x, 4, 1.5)
)

density_df <- tibble(
  x = density_estim$x,
  True = density_true,
  Estimation = density_estim$y
) |> 
  pivot_longer(cols = c(True, Estimation))

ggplot(density_df) +
  geom_line(aes(x = x, y = value, color = name)) +
  xlab("") +
  ylab("Density") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.9, 0.9)
  )
```

:::


::: {.callout-note appearance="simple"}

## Example: Simulation of a Zero-Inflated model.

Zero-inflated models are commonly used to model count data. The distribution of counts if often represented using a Poisson distribution. In some data, the number of zeros is greater than would be expected using a Poisson distribution. These data are referred as zero-inflated and can be simply described using a mixture model.

We aim to generate a sample $x_1, \dots, x_n$ i.i.d. from 
$$f(x) = 0.5\delta_0(x) + 0.5f_1(x),$$
where $\delta_0$ is the function that returns $0$ and $f_1$ is the probability mass function of $\mathcal{P}(5)$.

```{r}
#| class-output: outputcode
#| eval: true
set.seed(42)  # seed setting
n <- 10000  # number of samples 
probs <- c(0.5, 0.5)  # probabilities to be in each mixture

# Generate the pi variable and each mixture separetely
pi <- sample(c(1, 2), n, replace = TRUE, prob = probs)
f1 <- rpois(n, 5)

# Create the mixture
f <- (pi == 1) * 0 + (pi == 2) * f1

```

```{r}
#| class-output: outputcode
#| eval: true
# Compute the true probabilities and an estimation
prob_estim <- as.vector(table(f) / n)[1:15]
prob_true <- 0.5 + 0.5 * exp(-5)
for (i in 1:14) {
    prob_true <- c(prob_true, 0.5 * 5^i * exp(-5) / factorial(i))
}

probs_df <- tibble(
    x = seq(0, 14, by = 1),
    True = prob_true,
    Estimation = prob_estim
) |> 
    pivot_longer(cols = c(True, Estimation))

ggplot(probs_df) +
    geom_col(
        aes(x = x, y = value, fill = name),
        position = "dodge"
    ) +
    xlab("") +
    ylab("Probability") +
    theme_bw() +
    theme(
        legend.title = element_blank(),
        legend.position = c(0.9, 0.9)
    )

```

:::

## Additional resources

* The R package [mixtools](https://cran.r-project.org/web/packages/mixtools/index.html).

<br><br>

::: {style="font-size: 0.875em;"}
[Back](/weeks/week-8.qmd) ⏎
:::