---
title: "Simulation of linear models"
engine: knitr
---

```{r}
#| eval: true
#| include: false
#| warning: false
#| class-output: outputcode
library(tidyverse)
```

<br>
Simulating linear models is a straightforward process. We generate artificial data using known model parameters and random noise. By specifying the model equation, coefficients, and error structure, we can create datasets that mimic real-world scenarios. This approach is valuable for testing statistical techniques, evaluating model assumptions, and understanding the behavior of linear models under different conditions.

This section is adapted from the R Programming for Data Science [book](https://bookdown.org/rdpeng/rprogdatascience/simulation.html#simulating-a-linear-model).

## A simple linear model

Suppose that we are interested by the relationship between height and circumference of a tree. We would like to model the height as a function of the circumference and some random noise:
$$\text{height} = f(\text{circumference}) + \epsilon$$

We may assume that the function $f$ is linear. We have then the following model:
$$y = \beta_0 + \beta_1 x + \epsilon,$$

where $\epsilon \sim \mathcal{N}(0, 1)$. We assume that the circumference of a tree is uniform between $30$ and $70$ centimeters. We fix $\beta_0 = 9$ (the height if the circumference is equal to $0$) and $\beta_1 = 0.25$ (when the circumference of the tree grows by $1$cm, its height grows by $0.25$).

```{r}
#| class-output: outputcode
#| eval: true
set.seed(42)
n <- 100
x <- runif(n, 30, 70)
eps <- rnorm(n, 0, 1)
y <- 9 + 0.25 * x + eps

ggplot() +
  geom_point(aes(x = x, y = y)) +
  xlab("Circumference") +
  ylab("Height") +
  theme_bw()

```

We can simulate more complex models by adding more features (`x`) or their transformations (`x^2`, `log(x)`, ...), changing the distribution of the features or of the noise.

## A count linear model

As we previously saw, count data are often modelled with a Poisson distribution. We will generate data for this type of model. Let's generate data using the following model:
$$Y \sim \mathcal{P}(\mu) \quad \text{and}\quad \log\mu = \beta_0 + \beta_1 x + \epsilon,$$
where $\epsilon = \mathcal{N}(0, 0.0625)$. We fix $\beta_0 = 0.5$ and $\beta_1 = 0.3$.


```{r}
#| class-output: outputcode
#| eval: true
set.seed(42)
n <- 1000
x <- runif(n, -1, 1)
eps <- rnorm(n, 0, 0.25)
mu <- 0.5 + 0.3 * x + eps
y <- rpois(n, exp(mu))

ggplot() +
  geom_point(aes(x = x, y = y)) +
  xlab("") +
  ylab("") +
  theme_bw()
```

We can fit a Poisson log-linear model to the generate data to see if we are able to retrieve the coefficients we fixed. We use the `glm` function in R to do so. The results seem pretty good!

```{r}
#| class-output: outputcode
#| eval: true
glm(y ~ x, family = poisson(link = "log"))
```


## Additional resources

* R Programming for Data Science [book](https://bookdown.org/rdpeng/rprogdatascience/).

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 08](/weeks/week-8.qmd) ⏎
:::