---
title: "Interactive graphics with plotly"
engine: knitr
---

<br>
Interactive graphics enable users to explore data and gain insights more dynamically. This interactivity includes zooming, panning, hovering over data points for additional information, and selecting specific elements for deeper exploration. Nowadays, there are numerous R packages to create interactive plots. Here, we will focus on `plotly` as it is a general purposes data visualisation package and it works nicely with `ggplot2`.

Before you start, you will need to download and install the `plotly` package. We will also need the `tidyverse`.

```{r}
#| eval: false
#| class-output: outputcode
install.packages('plotly')
```

```{r}
#| eval: true
#| warning: false
#| class-output: outputcode
library(tidyverse)
library(plotly)
```

For this section, we will use a dataset gathering information on US States ([link](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-08-01)). The dataset is assumed to be loaded using:

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
df <- read_csv('./source/states.csv')
head(df)
```

## Getting started

The plotting philosophy of `plotly` is similar to the one of `ggplot2` in the sense that the plot is created by adding **layers** of information. The plots created with `plotly` are highly customisable.

The figures are created using the `plot_ly()` function and the layout can be modified using the `layout()` function. If we assign variable names (e.g. `state`, `population_2020`, etc) to visual properties (e.g. `x`, `y`, `color`, etc), `plot_ly()` will try to create a sensible plot based on the information we provide. The behaviour can be changed by setting the `type` parameters.

::: {.callout-note appearance="simple"}

Note that to assign a variable name to a visual property, we have to use the `~` symbol. For example, `x = ~population_2020` or `color = ~n_representatives`.

:::

::: {.callout-note appearance="simple"}

## Example

We want a plot for the `n_representatives` variable and `plot_ly()` choose to do a bar plot.

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
plot_ly(data = df, x = ~n_representatives)

```

:::

The `plotly` package considers a functional approach to plotting, in the sense, that most of the functions in this package take a `plotly` object as argument and return a modified version of this object. We use the `plot_ly()` function to define the plot and use the `add_*()` functions (e.g., `add_lines()`, `add_trace()`, ...) that define how the data is rendered.

It is important to note that:

* Arguments specified in `plot_ly()` are **global**, so any downstream `add_*()` functions inherit theses arguments, except if `inherit = FALSE`.

* Data manipulation verbs from the `dplyr` package may be used to transform the data underlying a `plotly` object. We use the `plotly_data()` to obtain the data at any point in the code.

::: {.callout-note appearance="simple"}

## Example

We do the same plot as before, but this time, we add the number of states with a certain number of representatives. This is done directly in the plot definition without the need to create a new dataframe.

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
plot_ly(
    data = df,
    x = ~factor(n_representatives)  # Globally assign x
) |> 
    add_histogram() |>  # Add a histogram layer
    group_by(n_representatives) |> 
    summarise(n = n()) |>  # Modify the data in the plotly object
    add_text(  # Add a layer of text, x is the same as before
        text = ~n, y = ~n, 
        textposition = "top middle", 
        cliponaxis = FALSE
    )
```

:::


The plots created with `plot_ly()` are very customisable. The `layout()` function is used to add and modify part of the graphics. For example, we can change the title, axis, grid or legend using this function. It works simularly to the `theme()` function in `ggplot2`. We select the element we want to modify (e.g., the `xaxis`) and we create a list of the characteristics we desire (e.g., `list(title = 'Population', color = 'red')`).


::: {.callout-note appearance="simple"}

## Example

A pretty complete example to create a scatter plot in `plotly`. 

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
plot_ly(
    data = df,
    type = 'scatter',
    mode = 'markers',
    x = ~population_2020,
    y = ~total_area_mi2,
    text = ~paste(
        'Population:', population_2020,
        '\nTotal area in mi^2:', total_area_mi2,
        '\nNumber of representative:', n_representatives
    ),
    marker = list(
        color = ~n_representatives, 
        colorbar = list(
            title = "Number of representatives",
            orientation = 'h'
        ),
        colorscale = 'Viridis',
        showscale = TRUE
    )
) |> 
    layout(
        title = 'Population vs Total Area of American States (in 2020)',
        xaxis = list(title = 'Population', color = 'red'),
        yaxis = list(title = "Total area in miles squared")
    )
```

:::



## Make a ggplot interactive

Making a `ggplot` interactive is very easy. The package `plotly` provides the function `ggplotly()` that does all the job. The function takes a `ggplot` object as input and return a `plotly` object. As it returns a `plotly` object, we can use all the formatting elements on it.

::: {.callout-note appearance="simple"}

## Example

We can create a simple plot using `ggplot2`:
```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
gg <- ggplot(df) +
    geom_point(
        aes(x = population_2020, y = total_area_mi2, text = state)
    ) +
    xlab('Population in 2020') +
    ylab('Total area in square miles') +
    theme_bw()
gg
```

And then, make it interactive using the `ggplotly()` function.

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
ggplotly(gg, tooltip = 'text')
```
:::


## Configuration

The `plotly` package also includes a `config()` function that can be used to set the configuration options of the figures (by default, the top-right buttons, named the modebar). For example, we can enable the scroll zoom by setting `scrollZoom = TRUE` or make a static chart with `staticPlot = TRUE`. 

Most of the configuration options are [here](https://plotly.com/r/configuration-options/).

## Additional resources

* Plotly R [documentation](https://plotly.com/r/).

* Plotly with `ggplot2` [documentation](https://plotly.com/ggplot2/).

* Interactive web-based data visualization with R, plotly and shiny [book](https://plotly-r.com).

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 06](/weeks/week-6.qmd) ⏎
:::