---
title: "Interactive graphics with plotly"
engine: knitr
---

<br>
Interactive graphics enable users to explore data and gain insights more dynamically. This interactivity includes zooming, panning, hovering over data points for additional information, and selecting specific elements for deeper exploration. Nowadays, there are numerous R packages to create interactive plots. Here, we will focus on `plotly` as it is a general purposes data visualisation package and it works nicely with `ggplot2`.

Before you start, you will need to download and install the `plotly` package. We will also need the `tidyverse`.

```{r}
#| eval: false
#| class-output: outputcode
install.packages('plotly')
```

```{r}
#| eval: true
#| warning: false
#| class-output: outputcode
library(tidyverse)
library(plotly)
```

For this section, we will use a dataset gathering information on US States ([link](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-08-01)). The dataset is assumed to be loaded using:

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
df <- read_csv('./source/states.csv')
head(df)
```

## Make a ggplot interactive

Making a `ggplot` interactive is very easy. The package `plotly` provides the function `ggplotly()` that does all the job. The function takes a `ggplot` object as input and return a `plotly` object. As it returns a `plotly` object, we can use all the formatting elements on it.

::: {.callout-note appearance="simple"}

## Example

We can create a simple plot using `ggplot2`:
```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
gg <- ggplot(df) +
    geom_point(
        aes(x = population_2020, y = total_area_mi2, text = state)
    ) +
    xlab('Population in 2020') +
    ylab('Total area in square miles') +
    theme_bw()
gg
```

And then, make it interactive using the `ggplotly()` function.

```{r}
#| class-output: outputcode
#| eval: true
#| warning: false
ggplotly(gg, tooltip = 'text')
```
:::


## Additional resources

* Plotly R [documentation](https://plotly.com/r/).

* Plotly with `ggplot2` [documentation](https://plotly.com/ggplot2/).

<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 06](/weeks/week-6.qmd) ⏎
:::