---
title: "Debugging and profiling - Tutorial"
engine: knitr
---

<br>
The aim of today's tutorial is to find and fix bugs in R functions and do some profiling of common R functions to see if some optimisation is feasible.


For the following code snippets, find and fix the bugs using `browser()` and `debug()`. It may have multiple bugs in each snippets. In a second step, do the profiling of the snippets using `profvis` and `bench`. You can try the functions with different inputs and try to understand if they can be optimised in some way.


```{r}
#| eval: false
calc_factorial <- function(n) {
    result <- 0
    for (i in 1:n) {
        result <- c(result, result * i)
    }
    return(result[length(result)])
}
```

```{r}
#| eval: false
is_prime <- function(n) {
  if (n <= 1) {
    return(FALSE)
  }
  if (n %% 2 == 0 || n %% 3 == 0) {
    return(FALSE)
  }
  i <- 5
  while (i * i <= n) {
    if (n %% i == 0 || n %% (i + 2) == 0) {
      return(FALSE)
    }
    i <- i + 4
  }
  return(TRUE)
}

sum_primes <- function(limit) {
  if (limit < 2) {
    return(0)
  }
  primes <- c(2)
  sum <- 2

  for (i in 3:limit) {
    if (is_prime(i)) {
      primes <- c(primes, i)
      sum <- sum + i
    }
  }
  
  return(list("primes" = primes, "sum" = sum))
}
```


<br><br>

::: {style="font-size: 0.875em;"}
Back to [week 09](/weeks/week-9.qmd) ⏎
:::